---
title: Resident orcas
editor: 
  markdown: 
    wrap: 72
---

Resident orcas simulate the Southern Resident Killer Whale endemic
population

-   Updated the diet using Ford et al. 1998, Ford & Ellis, 2006, Ford
    2009, Hanson et al., 2010 & 2021 and Hermann et al. 2015
    
```{r}

source("/home/atlantis/psatlantismodelupdates/inputs/libraries.R")
#External drive (currently G:) /Azure_download/ dietps
#Original data in Diet_database_V1_no_figures.xlsx
#Diet data estimated using Dirichelet in Diet_PS_rev2023.R
#Generated MLE of diet contributions
#diet matrix sums proportions to 1, created to illustrate diets

func.groups <- readr::read_csv(here::here("inputs","PugetSoundAtlantisFunctionalGroups.csv")) %>% 
  dplyr::mutate(prey = Code)

diet.matrix <- data.table::fread(here::here("inputs","ps_diet_matrix_Dec_2023.csv"))

prey.order <- c("Pacific Herring Puget Sound","Pacific Herring Cherry Point",
                "Chinook Hatch Subyearling","Chinook Hatch Yearling",  "Chinook other Subyearling", "Chinook Snohomish Subyearling","Chinook Skagit Subyearling","Chinook Skagit Yearling","Chinook other Yearling",            
                "Chinook Duwamish Subyearling", "Chinook Hood Canal Subyearling", "Chinook Nisqually Subyearling","Chinook Nisqually Yearling",
                "Chum Fall Subyearling", "Chum Hood Canal summer run Subyearling","Chum Hatch Subyearling",  
                "Coho Deep South Yearling", "Coho Hatch Yearling","Coho Skagit Yearling", "Coho other Yearling",    
                "Pink Salmon Subyearling", "Strait of Georgia salmonids", "Other salmonids",
                "Large demersal predators", "Small demersal fish",  "Small-mouthed Flatfish", "Piscivorous flatfish",  
                "Demersal rockfish", "Midwater rockfish",   "Demersal vulnerable rockfish",        
                "Skates",  
                "Squid")  



fill <- c("violet","violetred3",
          "lightsalmon","lightsalmon1","lightsalmon2","lightsalmon3","lightsalmon4","salmon","salmon1","salmon2","salmon3","salmon4","lightcoral",
          "rosybrown1","rosybrown2","rosybrown3",
          "indianred","indianred1","indianred2","indianred3",
          "sienna1","sienna2","sienna3",
          "deepskyblue4","dodgerblue4","navyblue","lightsteelblue3",
          "red1","red2","red4",
          "turquoise",
          "turquoise4"
          
)


diet.matrix.ROR <- diet.matrix %>% 
  dplyr::filter(predator=="Resident Orca") %>% 
  tidyr::pivot_longer(cols=(BB:DC_s), names_to="prey") %>% 
  dplyr::filter(value!=0) %>% 
  droplevels() %>% 
  dplyr::left_join(func.groups, by="prey") %>% 
  dplyr::mutate(prey = as.factor(`Long Name`)) 

diet.matrix.ROR$prey <- factor(diet.matrix.ROR$prey, levels = prey.order)

ror.diet.plot <- diet.matrix.ROR %>%  
  dplyr::filter(pred_stage=="adult") %>% 
  ggplot2::ggplot(ggplot2::aes(x=pred_stage, y=value, fill = prey)) + 
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::coord_flip() +
  scale_fill_manual(values=fill)+
  labs(x = "Resident orcas (adult)",y="Proportion") +
  #facet_wrap(~prey)+
  theme_bw()+
  theme(legend.position="bottom",
        legend.title = element_blank())

invert.data %>% 
  group_by(Predator, prey) %>% 
  summarise(mean_prop = mean(proportion)) %>% 
  ungroup %>% 
  left_join(invert.tot.prop, by="Predator") %>% 
  mutate(norm_prop = mean_prop/tot_prop) %>% 
  ggplot(aes(Predator, norm_prop, fill=prey))+
  geom_bar(stat="identity")+
  coord_flip()+
  scale_fill_manual(values=fill)+
  labs(x = "Predator",y="Proportion")+
  #  facet_wrap(~Predator)+
  theme_bw()+
  theme(legend.position="bottom")


```
    

-   Updated diet proportions for Chinook populations from Hanson et al.
    2021

-   Spatial distributions based on SRKW sightings. Estimated based on
    Thorson et al. 2015

``` {r}
    install.packages("sdmTMB", dependencies = TRUE)

```

-   Seasonal distribution

-   Fecundity. Fishery Management Council Working Group, data from Eric
    Ward, https://nwfsc-cb.github.io/srkw-status/.

**References**

Ford, J.K.B., G.A. Ellis, L.G. Barrett-Lennard, A.B. Morton, R.S. Palm &
K.C. Balcomb. 1998. Dietary specialization in two sympatric populations
of killer whales (*Orcinus orca)* in coastal British Columbia and
adjacent waters. Canadian Journal of Zoology-Revue Canadienne de
Zoologie 76: 1456--1471.

Ford, J.K.B. & G.M. Ellis. 2006. Selective foraging by fish-eating
killer whales *Orcinus orca* in British Columbia. Mar. Ecol. Prog. Ser.
316: 185--199.

Ford, J.K.B. 2009. Chinook salmon predation by resident killer whales:
Seasonal and regional selectivity, stock identity of prey, and
consumption rates. Canadian Science Advisory Secretariat = Secrétariat
canadien de consultation scientifique.

Thorson, J.T., Shelton, A.O., Ward, E.J., Skaug, H.J., 2015.
Geostatistical delta-generalized linear mixed models improve precision
for estimated abundance indices for West Coast groundfishes. ICES J.
Mar. Sci. J. Cons. 72(5), 1297--1310. doi:10.1093/icesjms/fsu243. URL:
<http://icesjms.oxfordjournals.org/content/72/5/1297>
