---
title: Resident orcas
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_packages
#| message: false
#| warning: false
#| include: false

library(readr)
library(ggplot2)
library(data.table)
library(dplyr)
library(tidyr)
```
Resident orcas simulate the Southern Resident Killer Whale endemic
population

-   Updated the diet using Ford et al. 1998, Ford & Ellis, 2006, Ford 2009, Hanson et al., 2010 & 2021 and Hermann et al. 2015
    
```{r}
#| label: diet_plots
#| echo: false
source(here::here("libraries.R")) #need to figure out which ones are actually needed
#External drive (currently G:) /Azure_download/ dietps
#Original data in Diet_database_V1_no_figures.xlsx
#Diet data estimated using Dirichelet in Diet_PS_rev2023.R
#Generated MLE of diet contributions
#diet matrix sums proportions to 1, created to illustrate diets

func.groups <- readr::read_csv(here::here("inputs","PugetSoundAtlantisFunctionalGroups.csv")) %>% 
  dplyr::mutate(prey = Code)

diet.matrix <- data.table::fread(here::here("inputs","ps_diet_matrix_Dec_2023.csv"))

prey.order <- c("Pacific Herring Puget Sound","Pacific Herring Cherry Point",
                "Chinook Hatch Subyearling","Chinook Hatch Yearling",  "Chinook other Subyearling", "Chinook Snohomish Subyearling","Chinook Skagit Subyearling","Chinook Skagit Yearling","Chinook other Yearling",            
                "Chinook Duwamish Subyearling", "Chinook Hood Canal Subyearling", "Chinook Nisqually Subyearling","Chinook Nisqually Yearling",
                "Chum Fall Subyearling", "Chum Hood Canal summer run Subyearling","Chum Hatch Subyearling",  
                "Coho Deep South Yearling", "Coho Hatch Yearling","Coho Skagit Yearling", "Coho other Yearling",    
                "Pink Salmon Subyearling", "Strait of Georgia salmonids", "Other salmonids",
                "Large demersal predators", "Small demersal fish",  "Small-mouthed Flatfish", "Piscivorous flatfish",  
                "Demersal rockfish", "Midwater rockfish",   "Demersal vulnerable rockfish",        
                "Skates",  
                "Squid")  



fill <- c("violet","violetred3",
          "lightsalmon","lightsalmon1","lightsalmon2","lightsalmon3","lightsalmon4","salmon","salmon1","salmon2","salmon3","salmon4","lightcoral",
          "rosybrown1","rosybrown2","rosybrown3",
          "indianred","indianred1","indianred2","indianred3",
          "sienna1","sienna2","sienna3",
          "deepskyblue4","dodgerblue4","navyblue","lightsteelblue3",
          "red1","red2","red4",
          "turquoise",
          "turquoise4"
          
)


diet.matrix.ROR <- diet.matrix %>% 
  dplyr::filter(predator=="Resident Orca") %>% 
  tidyr::pivot_longer(cols=(BB:DC_s), names_to="prey") %>% 
  dplyr::filter(value!=0) %>% 
  droplevels() %>% 
  dplyr::left_join(func.groups, by="prey") %>% 
  dplyr::mutate(prey = as.factor(`Long Name`)) 

diet.matrix.ROR$prey <- factor(diet.matrix.ROR$prey, levels = prey.order)

ror.diet.plot <- diet.matrix.ROR %>%  
  dplyr::filter(pred_stage=="adult") %>% 
  ggplot2::ggplot(ggplot2::aes(x=pred_stage, y=value, fill = prey)) + 
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(values=fill)+
  ggplot2::labs(x = "Resident orcas (adult)",y="Proportion") +
  #facet_wrap(~prey)+
  ggplot2::theme_bw()+
  ggplot2::theme(legend.position="bottom",
        legend.title = element_blank())

ror.diet.plot


```
    

-   Updated diet proportions for Chinook populations from Hanson et al. 2021

-   Spatial distributions based on SRKW sightings. Estimated based on Thorson et al. 2015

-   Seasonal distribution

We updated spatial distribution based on orca master data of killer whale sightings for the years 2011 - 2020. Laura Kohen (NOAA-Western Regional Center, Protected Species Division) carried out the following analysis: To match sightings of Southern Resident killer whales to Puget Sound Atlantis boxes. First, sightings are filtered to only sightings where there was confidence that the sighting was of Southern Resident killer whales (vs other killer whale populations). Specifically, data is filtered to sightings where the "likely pod" is either SRKW, SRs, J, K, or L; so excludes any sightings where the sighting could only be identified to the "orca" level (not more specific). Also excludes sightings that were specified as transients, Northern residents, "not orcas", or HBs (humpbacks). Data was then filtered to years 2011-2020. From there, we filtered down to unique day-quadrant sightings. Specifically, if there are any 2+ sightings with the same date and Whale Museum specified quadrant number, we only keep 1 sighting with that day-quadrant combo. This is to remove sightings from multiple sources that can inflate occurrence in an area. Note, this then means that there could be multiple sightings on the same day in the same Atlantis box, but they will all be in different Whale Museum quadrants. 

Note also that data was filtered down to only sightings with a whale museum identified quadrant number because these will occur in the Salish sea. Note that Whale Museum quadrant extend passed the Atlantis box domain (we do not have a shapefile of the quadrants but image is included with data). This may miss a few sightings if a quadrant was not specified by the Whale Museum but not a large amount. 

With the sightings data, sightings were converted to an sf object with latitude and longitude. Then, based on the CRS of the Atlantis polygons, the SRKW sightings sf were projected to the same CRS as the Atlantis polygons so that all are in the same sf space/same traget coordinate reference system (function st_transform). Then using st_intersection, this function finds the Atlantis polygons that each sighting falls into. 

The final output file has for all the day-quadrant unique SRKW sightings from 2011-2020, the date (including month, day, year) and the Atlantis polygon where the sighting fell. 


-   Fecundity. Fishery Management Council Working Group, data from Eric Ward, https://nwfsc-cb.github.io/srkw-status/.

**References**

Ford, J.K.B., G.A. Ellis, L.G. Barrett-Lennard, A.B. Morton, R.S. Palm & K.C. Balcomb. 1998. Dietary specialization in two sympatric populations of killer whales (*Orcinus orca)* in coastal British Columbia and adjacent waters. Canadian Journal of Zoology-Revue Canadienne de Zoologie 76: 1456--1471.

Ford, J.K.B. & G.M. Ellis. 2006. Selective foraging by fish-eating killer whales *Orcinus orca* in British Columbia. Mar. Ecol. Prog. Ser. 316: 185--199.

Ford, J.K.B. 2009. Chinook salmon predation by resident killer whales: Seasonal and regional selectivity, stock identity of prey, and consumption rates. Canadian Science Advisory Secretariat = Secrétariat canadien de consultation scientifique.

Thorson, J.T., Shelton, A.O., Ward, E.J., Skaug, H.J., 2015. Geostatistical delta-generalized linear mixed models improve precision for estimated abundance indices for West Coast groundfishes. ICES J. Mar. Sci. J. Cons. 72(5), 1297--1310. doi:10.1093/icesjms/fsu243. URL: <http://icesjms.oxfordjournals.org/content/72/5/1297>
